CursedForest - A Random Forest Implementation for ``Big'' and ``Wide'' Data

### 2017-05-19 doc/publications/IJCAI2017/IJCAI.tex
added conference paper

% Template for PLoS
% Version 3.4 January 2017
%
% % % % % % % % % % % % % % % % % % % % % %
%
% -- IMPORTANT NOTE
%
% This template contains comments intended 
% to minimize problems and delays during our production 
% process. Please follow the template instructions
% whenever possible.
%
% % % % % % % % % % % % % % % % % % % % % % % 
%
% Once your paper is accepted for publication, 
% PLEASE REMOVE ALL TRACKED CHANGES in this file 
% and leave only the final text of your manuscript. 
% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
%
%
% There are no restrictions on package use within the LaTeX files except that 
% no packages listed in the template may be deleted.
%
% Please do not include colors or graphics in the text.
%
% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
%
% % % % % % % % % % % % % % % % % % % % % % %
%
% -- FIGURES AND TABLES
%
% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
%
% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
% - Figures should be uploaded separately from your manuscript file. 
% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
% For figure citations, please use "Fig" instead of "Figure".
% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
%
% Tables should be cell-based and may not contain:
% - spacing/line breaks within cells to alter layout or alignment
% - do not nest tabular environments (no tabular environments within tabular environments)
% - no graphics or colored text (cell background color/shading OK)
% See http://journals.plos.org/plosone/s/tables for table guidelines.
%
% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
%
% % % % % % % % % % % % % % % % % % % % % % % %
%
% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
%
% IMPORTANT
% Below are a few tips to help format your equations and other special characters according to our specifications. For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
%
% For inline equations, please be sure to include all portions of an equation in the math environment.  For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
%
% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
%
% Please add line breaks to long display equations when possible in order to fit size of the column. 
%
% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
%
% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
%
% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
%
% % % % % % % % % % % % % % % % % % % % % % % % 
%
% Please contact latex@plos.org with any questions.
%
% % % % % % % % % % % % % % % % % % % % % % % %

\documentclass[10pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% color can be used to apply background shading to table cells only
\usepackage[table]{xcolor}

% array package and thick rules for tables
\usepackage{array}

% create "+" rule type for thick vertical lines
\newcolumntype{+}{!{\vrule width 2pt}}

% create \thickcline for thick horizontal lines of variable length
\newlength\savedwidth
\newcommand\thickcline[1]{%
  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
  \cline{#1}%
  \noalign{\vskip\arrayrulewidth}%
  \noalign{\global\arrayrulewidth\savedwidth}%
}

% \thickhline command for thick horizontal lines that span the table
\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
\hline
\noalign{\global\arrayrulewidth\savedwidth}}

% Remove comment for double spacing
%\usepackage{setspace} 
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in 
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\renewcommand{\figurename}{Fig}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2015}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{epstopdf}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\setlength{\headheight}{27.023pt}
\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

%% END MACROS SECTION

%% Additional packages and local edits
\usepackage{footmisc}
\usepackage{multirow}
\usepackage[T1]{fontenc}
%% required the installation of cm-super 55MB - TeX font package (full version) with CM (EC) in Type1 in T1, T2*, TS1, X2 enc

\let\oldmarginpar\marginpar
\renewcommand\marginpar[1]{\-\oldmarginpar[\raggedleft\footnotesize #1]%hv
{\raggedright\footnotesize #1}}
\reversemarginpar

%\setlength{\marginparwidth}{5cm}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{placeins}
\usepackage{xspace} % see http:
%\usepackage{showframe}

@@

\begin{document}
\vspace*{0.2in}

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{\cursedforest - A Random Forest Implementation for ``Big'' and ``Wide'' Data} % Please use "title case" (capitalize all terms in the title except conjunctions, prepositions, and articles).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Aidan O'Brien\textsuperscript{1,2},
Piotr Szul\textsuperscript{3},
Robert Dunne\textsuperscript{4}, 
Paul Leo \textsuperscript{5},
Emma L. Duncan \textsuperscript{5,6,7},
Natalie A. Twine\textsuperscript{1,8}
Stephanie Li\textsuperscript{9},
James Doecke\textsuperscript{9},
Nick Ellis\textsuperscript{10}, and
Denis C. Bauer\textsuperscript{1,*}
%, with the Lorem Ipsum Consortium\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Health \& Biosecurity, CSIRO, Sydney, NSW, Australia
\\
\textbf{2} ANU, Canberra, Australia
\\
\textbf{3} Data61, CSIRO, Brisbane, QLD, Australia
\\
\textbf{4} Data61, CSIRO, Sydney, NSW, Australia
\\
\textbf{5} Institute of Health and Biomedical Innovation, Queensland University of Technology, Brisbane, QLD, Australia
\\
\textbf{6} Department of Endocrinology, Royal Brisbane and Women's Hospital, Brisbane, QLD, Australia
\\
\textbf{7} Faculty of Medicine and Biomedical Sciences, University of Queensland, Brisbane, QLD, Australia
\\
\textbf{8} MQ, QLD, Australia
\\
\textbf{9} Health \& Biosecurity, CSIRO, Brisbane, QLD, Australia
\\
\textbf{10} Oceans \& Atmosphere, CSIRO, Brisbane, QLD, Australia
\bigskip

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
% 
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Membership list can be found in the Acknowledgments section.

% Use the asterisk to denote corresponding authorship and provide email address in note below.
* Denis.Bauer@CSIRO.au
\end{flushleft}

%\tableofcontents
%\clearpage

% Please keep the abstract below 300 words
\section{Abstract}

Analyses are performed on ever larger datasets with unprecedentedly detailed information per data point. Specifically in
genomics, cohort sizes are in the thousands, each comprising of millions of variants that make up a person's genetic
differences. Traditional approaches are unable to cater for this task and even standard implementations using novel
compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory on such
data. Specifically, Spark ML's supervised machine learning methods suffer from "the curse of dimensionality". That is,
they scale well with samples ($n$) but not with features ($p$, variants).

We introduce \cursedforest, a tailored implementation of random forests, designed to handle data with extremely large
number of variables per sample. We showcase \cursedforest\ to accurately predict ethnicity from whole genome sequencing
profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. \cursedforest\
was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or
more. We also demonstrate a GWAS-style analysis using \cursedforest's feature selection approach on an imputed SNP array
dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying
novel candidates.

\cursedforest is included in our earlier variant interpretation package, VariantSpark, allowing it to perform near
real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing
GWAS analysis on large unfiltered whole genome sequencing cohorts.

\linenumbers

@@

The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \randomforest implementation in R lending itself to parallelisation by growing multiple
trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
multithreading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
specifically optimised for dealing with large datasets.

@@

Here we introduce \cursedforest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) datasets. 
\cursedforest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
supervised as well as unsupervised ML algorithms in the Spark framework.
In our implementation a Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
directly, \cursedforest outperforms existing tools even on small datasets where multithreading generally performs
well. Harnessing the virtually unlimited capability to parallelise tasks, \cursedforest can hence explore the solution 
space faster by building a larger number of diverse models to generate a consensus from.

Using this facility, \cursedforest is capable of parallelising the split for each node in a tree thereby handling millions 
of features, as required to process whole genome sequencing data or SNP array data with unobserved genotypes 
imputed~\cite{Howie2012}.  This provides the potential to generate datasets of hundreds of thousands of individuals 
with millions of variants (imputing the GWAS catalog), highlighting the need for modern
compute paradigms in the genomics space.

%It can, for example, fit a model to the 1000 Genomes Project data, which consists of approximately 2500 samples with up to 80 million
%variants~\cite{1KG2012}.  
%build a larger number of
%trees hence building a larger number of diverse models to generate the consensus from.  %and use a high \mtry value which may lead to greater accuracy (see the example in section \ref{section:synthetic}). 
%\marginpar{does \cite{Szymczak2016} say that high \ntree gives greater  accuracy for variable importance selection?
%Check this. }
%We previously demonstrated the versatility and scalability of Spark by developing VariantSpark~\cite{OBrien2015}, a
%framework allowing users to easily analyze Variant Call Format (VCF) files using ML algorithms on the Spark framework.
%Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
%individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
%random clustering).
%Using VariantSpark, we successfully built a k-means model on the full $2500 \times 80$ million matrix to cluster
%individuals by their ethnicity achieving an Adjusted Rand Index (ARI) of 0.84 (with 1 being perfect clustering and -1
%random clustering).
%Here we extend this toolkit to encompass supervised learning using random forests.

VariantSpark~\cite{OBrien2015} with the \cursedforest extension therefore offers a
comprehensive analysis toolkit that can scale to future data demands. To showcase the framework's ability we demonstrate
a classification as well as feature-selection task on synthetic data in the first section. In the second section, we demonstrate the ability of
\cursedforest to successfully replicate findings from a previous GWAS study as well as identify novel variants associated with bone 
mineral density (BMD). 
Thirdly, we demonstrate the scalability of \cursedforest in respect to the dimensionality of data by building a random
forest model on whole-genome data from the 1000 Genomes Project~\cite{1KG2012} to predict ethnicity.
Finally, given the role different parameter values can play in model construction, we explore the effect that tuning these
parameters can have on the prediction accuracy of the model.

\section{Methods}
\label{section:methods}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Overview of Random Forests}

Decision trees have a number of desirable features, they:
\begin{itemize}
  \item are invariant to monotonic scaling of the data;
  \item can handle categorical and real valued inputs;
  \item can handle missing values;
  \item are insensitive to uninformative predictors;
  \item are able to capture interactions between features (which is important for modeling complex polygenic diseases).
  \end{itemize}
However, the tree fitting algorithm is greedy and may generate an unstable model. That is, a small change in the data may
lead to a very different model.

Random forests apply the technique of bootstrap aggregation or bagging to decision trees.  The training datasets are
independently drawn, with replacement, from the joint distribution of $(X,Y)$ so that each set comprises $n$
$(p+1)$-tuples $(x_1,y_1),\ldots, (x_n,y_n)$. This is done $B$ times and a tree is grown on each of these $B$ samples.
The results are then combined in an appropriate way. In a classification problem we take the majority vote for the trees
over the $q=1,..,Q$ classes,
\begin{equation*}
{{h_f}}= \arg \max_q \left(\sum_b I(h(x;B_b)=q)\right).
\end{equation*}
In addition to growing the trees
on bootstrapped samples, each node in the tree is split on a randomly
selected subset of the variables.

Random forests are a variance reduction technique. They take unpruned tree models, which have a low bias but a high
variance, and by combining them reduce the variance.  For regression problems, as the prediction error is the sum of the
variance and the bias squared, this strategy should lead to a low prediction error.

For classifiction problems there is no obvious decomposition of classification error into an equivalent bias plus
variance form. A number of different additive decompositions of classification error have been considered in an effort
to understand the success of bagging. They are all less transparent than the regression case (see \cite{Friedman.1997} and
citations therein).

In the case of a regression tree, the average prediction error for an individual tree $h(X; B)$ is
\begin{equation}
PE_t = E_B E_{X,Y} (Y-h(X; B))^2.
\end{equation}
Assume that, for all $B$, the tree is unbiased, i.e., $EY= E_X h(X; B)$. Then
\begin{equation}
PE_f \leq \rho PE_t
\end{equation}
where $\rho$ is the weighted correlation between residuals $(Y-h(X;B))$ and $(Y-h(X;B^\prime))$ for independent
$B,B^\prime$.
So what is required for accurate random forest regression is (i) low correlation between residuals of differing tree in
the forest, and (ii) low prediction error for the individual trees \cite{Segal.2004}.

The benefits of random forest models are:
\begin{itemize}
\item they give an out-of-bag (OOB) estimate of model accuracy, by using the samples not selected for the bootstrapped sample
  as an independent test set for each tree;
\item they provide a measure of variable importance, for example by summing the number of times a variable is used to
  split a node, over all nodes and all trees. More complex measures of variable importance are also used;
\item they inherit some of the qualities of tree models (but not, for example, the handling of missing values via surrogate
  variables. This would be difficult as the trees are split on random subsets of variables).
\end{itemize}

Since the original implementation of random forests, there have been a number of developments leading to a better
understanding of the algorithm.
\begin{itemize}
\item random forests are resistant to overfitting. However it is not true that they will not overfit,
  ~\cite{Segal.2004} gives examples of overfitting in cases where:
  \begin{itemize}
  \item a single tree is the most appropriate model; and
  \item the variables are highly correlated so that the bootstrap resampling does not give rise to uncorrelated trees.
  \end{itemize}
However, keeping these difficult cases in mind, random forests have a strong claim to being resistant to overfitting.
\item \cite{Strobl.et.al.2007} demonstrates that the random forest algorithm is subject to a bias in variable selection
  via two mechanisms:
  \begin{enumerate}
  \item differences in variable scale (for $X_i$ continuous) and number of levels (for $X_i$ discrete) introduces a
    bias. An uninformative variable with a large number of levels may be selected over a more informative variable with
    fewer levels;
  \item bagging introduces a bias, but sampling $(1- 1/e) \approx 0.632$ of the data without replacement seems to be an
    effective strategy for removing the problem.
  \end{enumerate}
\end{itemize}

\cite{Strobl.et.al.2007} introduce changes to the random forest algorithm to fix 1) and 2).  Effect 1) will be apparent
in the \cursedforest algorithm as well.  However, there are many examples, particularly in genomics, where we have data
sets that are:
\begin{itemize}
\item very wide;
\item all variables have the same number of levels (for example, measurements at different points along the genome);
\item the hypothesized functional relationship is of a form amenable to a decision tree.
\end{itemize}
GWAS datasets are an obvious application area for this algorithm.

\cite{Biau.2012} provides a proof that a random forest model will converge at a rate that depends on
the cardinality of the set $S$ of ``strong predictors'' rather than on the number of variables $p$. That is, given a
function $y=f(S)$ depending on a set of variables $S$, a random forest will converge to the true function $y$ with a
rate that depends on the size of $S$ and not the number of noise variables in the dataset. If the size of $S$ is small
compared to the total number of variables, then the rate of convergence will be quicker. It is not clear what the size
of the set $S$ will be in a GWAS application. There are cases of small groups of SNPs with large effects and also of
conditions caused by the interactions of larger groups of SNPs.

\cite{Genuer.et.al.2010} and \cite{Diaz.and.Alvarez.2006} describe iterative schemes for doing variable selection using
the importance measure. See \cite{Chen.and.Ishwaran.2012} for a review of the area with particular reference to genomic
data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest}
VariantSpark made use of Spark's machine learning algorithms ``Spark ML''. While Spark ML algorithms demonstrate
scalability when dealing with a large number of samples, this scalability eventually breaks down as we include more
features.

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can be distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.

@@

On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
work down to the level of the individual features.  For each node of a tree, \cursedforest will distribute
tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
entire decision tree.

The current implementation of \cursedforest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is at a minimum when all observations at the node are in the same class.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% This section talks about the datasets
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Datasets}
\subsubsection{1000 Genomes Project}
We obtained the 1000 Genomes Project data as VCF files from their FTP site.  Each VCF file contains the variants for
every individual across one chromosome.  We use the phase 3 dataset which contains 2,504 individuals with approximately
81 million variant positions (i.e. a matrix of $2,504 \times 81$ million).  For compatibility with the majority of the
algorithms we demonstrate, we convert the original variant representations to double representations.  For example, we
store the absence of a variant (0\textbar0) as 0, a heterozygous
variant (1\textbar0 or 0\textbar1) as 1, and a
homozygous variant (1\textbar1) as 2.  However, this is not the only possible encoding \cite{Goldstein.et.al.2011}. Note that from
the encodings listed in Table~\ref{table:encodings}, the additive encoding will present the most difficulty for a random
forest model.

\begin{table}[!ht]
%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\begin{flushleft} 
\centering
\caption{\bf Possible encodings of GWAS data \cite{Goldstein.et.al.2011}}
\begin{tabular}{|l|l|l|l||}
\hline
{\bf Type} & {\bf Mechanism}                                         & {\bf Partition} \\ 
\thickhline
Additive   & Each additional minor allele increases variation        & 0, 1, 2         \\
Dominant   & Presence of at least 1 minor allele increases variation & 0, $\{1,2\}$          \\
Recessive  & Two minor alleles needed for variation                  & $\{0,1\}$, 2          \\
Heterosis  & Heterozygote leads to variation                         & $\{0,2\}$, 1          \\ \hline
%\end{adjustwidth}
\end{tabular}
\label{table:encodings}
\end{flushleft}
\end{table}

\subsubsection{Bone mineral density dataset}
We obtained a VCF file from the authors of \cite{Duncan.2011}, which contains variant data from 1,936 postmenopausal women.
This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan.2011}.

\subsubsection{Synthetic data} 
\label{section:synthetic_data}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomic variation) randomly generated from a uniform distribution with equal
probabilities.

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the \randomforest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in \randomforest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that the question of reducing bias versus variance is more
    complicated than first thought. They show that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Results
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results and Discussion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2,500,000$), of which 5 
are related to the response variable, and 5000 samples ($n=5000$), as discussed in the methods section.

>>>>

\documentclass[a4paper]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{subcaption}

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

@@

\title{CursedForest - A Random Forest Implementation for ``Big'' and ``Wide'' Data}
%\section{}
%\subsection{}
\begin{document}  
\maketitle

\begin{abstract}
Genomic data analyses are performed on ever larger patient cohorts with unprecedentedly detailed genome-wide information for each individual. Machine learning (ML) is employed to detect the complex genomic interactions that can lead to diseases like diabetes or cancer. 
However, current ML approaches are unable to cope with these data volumes and even implementations using novel compute paradigms, such as Hadoop/Spark, are limited by the finite resource of computer memory. Specifically, Spark ML?s supervised machine learning methods suffer from "the curse of dimensionality". That is, they scale well with samples (n) but not with features (p, variants). 
We introduce CursedForest, a tailored implementation of random forests, designed to handle data with extremely large number of variables per sample. We showcase CursedForest to accurately predict ethnicity from whole genome sequencing profiles of 2,500 individuals and 80 million genomic variants each (1000 Genomes Project) in 30 minutes. CursedForest was the only implementation scaling to the full dataset, outperforming other methods on smaller subsets by 80\% or more. We also demonstrate a GWAS-style analysis using CursedForest?s feature selection approach on an imputed SNP array dataset (2000 x 7 Million) on Bone mineral density (BMD), successfully predicting known BMD-genes as well as identifying novel candidates. 
CursedForest is included in our earlier genome interpretation package, VariantSpark, allowing it to perform near real-time classification of population-scale patient cohorts in "patients like mine" scenarios, as well as performing GWAS analysis on large unfiltered whole genome sequencing cohorts. 
\end{abstract}

%\twocolumn

@@

The most popular random forest software packages to date are R-based, although it has been reported that current implementations 
do not scale well with increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the random forest implementation in R lending itself to parallelisation by growing multiple
trees simultaneously and combining them for the final result~\cite{Liaw.and.Weiner.2002}.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015}, written in Go,
achieves fast run times by effective use of the CPU cache, optimising for different classes of features and efficiently
multithreading.  Similarly, ranger~\cite{Wright.and.Ziegle.2016}, written in C++ with an R front-end, is
specifically optimised for dealing with large datasets.

@@

Here we introduce CursedForest, a tailored Hadoop/Spark-based implementation of random forests specifically designed to cater
for ``big'' (many samples) and ``wide'' (many features) datasets. 
CursedForest extends our previously developed variant interpretation framework, VariantSpark~\cite{OBrien2015}, to now offer
supervised as well as unsupervised ML algorithms in the Spark framework.
In our implementation a Spark application runs on a ``driver'' node and
distributes tasks to many ``worker'' nodes, or ``executors''. 
By also utilising Spark to read in and manipulate the standard genomic variant format (VCF)
directly, CursedForest outperforms existing tools even on small datasets where multithreading generally performs
well. Harnessing the virtually unlimited capability to parallelise tasks, CursedForest can hence explore the solution 
space faster by building a larger number of diverse models to generate a consensus from.

Using this facility, CursedForest is capable of parallelising the split for each node in a tree thereby handling millions 
of features, as required to process whole genome sequencing data or SNP array data with unobserved genotypes 
imputed~\cite{Howie2012}.  This provides the potential to generate datasets of hundreds of thousands of individuals 
with millions of variants (imputing the GWAS catalog), highlighting the need for modern
compute paradigms in the genomics space.

VariantSpark~\cite{OBrien2015} with the CursedForest extension therefore offers a
comprehensive analysis toolkit that can scale to future data demands. To showcase the framework's ability we demonstrate
a classification as well as feature-selection task on synthetic data in the first section. In the second section, we demonstrate the ability of
CursedForest to successfully replicate findings from a previous GWAS study as well as identify novel variants associated with bone 
mineral density (BMD). 
Thirdly, we demonstrate the scalability of CursedForest in respect to the dimensionality of data by building a random
forest model on whole-genome data from the 1000 Genomes Project~\cite{1KG2012} to predict ethnicity.
Finally, given the role different parameter values can play in model construction, we explore the effect that tuning these
parameters can have on the prediction accuracy of the model.

\section{Methods}
\subsection{CursedForest}
As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can be distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can simply schedule extra tasks to handle the additional items in the RDD.

@@

On the other hand, CursedForest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
work down to the level of the individual features.  For each node of a tree, CursedForest will distribute
tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
feature which gives the greatest information gain.  This process is then repeated until CursedForest has created the
entire decision tree.

The current implementation of CursedForest uses a ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labeled with value $q$, where $q = 1, \ldots, Q$ at a node. The Gini impurity is
$$
I_G(f) = \sum_{q = 1}^Q f_q ( 1 - f_q ), 
$$
which is at a minimum when all observations at the node are in the same class.

\subsection{Synthetic data} 
\label{section:synthetic_data}
 Each dataset consists of $n$ samples and $p$ variables where $n << p$,
and values for each variable are ordinal variables with three levels represented as numbers $\{0, 1, 2\}$ (which
correspond to an additive effect encoding of genomic variation) randomly generated from a uniform distribution with equal
probabilities.

@@

\subsection{Parameter settings}
We consider the parameter settings for the random forest algorithm. We use the R notation from the random forest
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item ntree  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item mtry  -- the number of variables considered at each split (if mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in random forest by setting \texttt{nodesize}, the
  minimum size of terminal nodes. It is controlled in Spark ML by setting \texttt{maxDepth}, the maximum depth of each
  tree in the forest.\footnote{The Spark ML documentation \cite{Spark.2016} sets \texttt{maxDepth} to 4 in their
    classification example. This may give an estimate with a higher bias and, as such, may be a poor choice. See the
    discussion in \cite{Dietterich.2002} where it is shown that the question of reducing bias versus variance is more
    complicated than first thought. They show that bagging may reduce bias as well as variance but give a
    better result for low bias learners. \cite{Dietterich.2002} notes that if the bootstrap replicate approximation were
    correct (i.e.~if the bootstrap sample came from an identical distribution to the data), then bagging would reduce
    variance without changing bias.}
\end{itemize}

\section{Results}
\subsection{Scalability}
In this section we explore the performance of CursedForest in more detail by testing its ability to scale to different sizes of data
and computational resources.

In order to asses these characteristics, we run CursedForest classification on synthetic datasets with varying numbers
of variables (features) and samples, similar to the dataset used in \cite{Wright.and.Ziegle.2016} to evaluate
ranger, allocating varying number of CPU cores to the CursedForest and also varying the computational complexity
of the random forests by using a range of mtry values.
